---
title: "An introduction to popcompr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to popcompr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`popcompr` is an R package to make it easier to compare different high resolution population datasets for humanitarian and research purposes. It is under active development and has not been released, but the code is available through a GPL3 license.

To install package from github: 
```{r, eval = FALSE}
remotes::install_github("mrajeev08/popcompr")
```


## An example comparing 2019 population estimates for Lesotho from World Pop & Facebook/CIESIN
Included in the package are files to reproduce a minimal working example for the country of Lesotho. `lesotho_wp_2019` and `lesotho_fb_2019` are included in the package (use `?lesotho_wp_2019` for more details).

### First load the necessary libraries 
+ some extras for plotting (plotly & ggplot2)
```{r setup}
library(popcompr)
library(raster)
library(data.table)
library(foreach)
library(plotly)
library(ggplot2)
library(fasterize)
library(sf)
```

### Comparing populations at pixel level

You can first get an estimate of the time required to return the comparison:
```{r}
# comparing at pixel level with data included in the package
lesotho_fb_2019 <- raster(system.file("external/lso_facebook_2019.tif", package="popcompr"))
lesotho_wp_2019 <- raster(system.file("external/lso_worldpop_2019.tif", package="popcompr"))

pop_list <- list(lesotho_wp_2019, lesotho_fb_2019)
compare_pop(pop_list, parallel = FALSE, 
                   estimate_time = TRUE)
```

It shouldn't take too long by that estimate, but here we'll work through parallelizing. Here, I use the 
doParallel backend, but other do packages can be used (anything compatible with the `%dopar%` infix in the `foreach` package).

```{r}
# parallelized example
library(doParallel)
cl <- makeCluster(detectCores() - 1) # how many cores do we have available
registerDoParallel(cl)
system.time(
  # defaults to estimate_time = FALSE & resolution of ~ 1km at equator
  exe <- compare_pop(pop_list, parallel = TRUE) 
)
stopCluster(cl)

```

`compare_pop` will warn you if any people were not resampled to the comparison grid (this happens sometimes when pops are at the edge of the original raster).

The output is a raster brick with a layer corresponding to each of the input population rasters. We can vizualize and compare the rasters:

#### Vizualizing and summarizing comparisons
`plot_compare` and `summary_compare` are convenience functions to generate automatic plots. 
The default is map of the differences between the population datasets.
```{r}
plot_compare(exe)
```

Other options include:
- hex plots
```{r}
plot_compare(exe, type = "hex")
```

- a pair plot (using function `GGally::ggpairs`)
```{r}
plot_compare(exe, type = "pairs")
```

- histogram
```{r}
plot_compare(exe, type = "hist")
```

All plots are ggplots and as such can be customized by adding scales, themes, etc. 
You may get a warning if you're overriding an existing theme.

```{r}
out <- plot_compare(exe, type = "hex")
out +
  scale_fill_distiller(palette = "Greens", trans = "log")
```

**Note that for the map plots, changing the scales will mess up the labeling of the values. 
This is known issue that I'll try to fix!**

You can also make these interactive using plotly:
```{r}
plot_compare(exe, type = "map", interactive = TRUE) 
```
**Note that for the tooltips for these plots show the transformed values, should
show the raw values, otherwise the interactive view is not that helpful!
This is known issue that I'll try to fix!**

If you're comparing a large number of rasters, you can also save the output, rather
than returning the plot to your R session (just pass a file path and an extension). 
```{r, eval = FALSE}
plot_compare(exe, type = "map", save = TRUE, path = "test", ext = "jpeg") 

# interactive plots get saved as html with package `htmlwidgets`
plot_compare(exe, type = "map", save = TRUE, interactive = TRUE,
             path = "test", ext = ".html") 

```

You can also generate summary stats. See `?summary_compare` for more details 
on how to customize the summary stats. 
```{r}
summary_compare(exe)
```

### Comparing at the administrative level
You can access country shapefiles using `get_country_shape` which will return a shapefile as an `sf` object. These are available through the [geoBoundaries API](https://www.geoboundaries.org/api.html), where more documentation can be found on available datasets. In addition, `geoboundaries` is a dataset provided in the package to find country codes and available admin levels. 
```{r}
# Find the right iso code & see which admin levels are available
dplyr::filter(geoboundaries, grepl("Les", country))
```

We get the Lesotho shapefile to the admin level 2 (the `type` argument defaults to a simplified shapefile which is faster for plotting and downloading, but users may prefer unsimplified files for better accuracy see the geoBoundaries API documentation for more information).
```{r}
les_shape <- get_country_shape(country_iso = "LSO", admin_level = 2)
```

Then we can aggregate the population rasters to it:

```{r}
# Get the shapefile at admin 2
les_shape <- aggregate_to_shp(brick = exe, sf = les_shape, max_adjacent = 100)
```
`aggregate_to_shp` will warn you if grid cell values go unallocated to the shapefile. In the case that many people go missing, you can set max_adjacent higher (this determines how many grid cells to buffer to when looking for the nearest non-NA neighbor), but it may also be wise to check the extent & boundaries of the shapefile vs. the population rasters.

We can compare these differences using the same functions as with the pixel 
level comparisons:
```{r}
plot_compare(les_shape)

plot_compare(les_shape, type = "hex")

plot_compare(les_shape, type = "pairs")

plot_compare(les_shape, type = "hist")
summary_compare(les_shape)
```

### User specific inputs
You can use your own raster and shapefile inputs, they just need to be in the WGS84 (lat/long) coordinate system.
