---
title: "An introduction to popcompr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to popcompr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`popcompr` is an R package to make it easier to compare different high resolution population datasets for humanitarian and research purposes. It is under active development and has not been released, but the code is available through a GPL3 license.

### An example comparing 2019 population estimates for Lesotho from World Pop & Facebook/CIESIN
Included in the package are files to reproduce a minimal working example for the country of Lesotho. `lesotho_wp_2019` and `lesotho_fb_2019` are included in the package (use `?lesotho_wp_2019` for more details).

#### First load the necessary libraries 
+ some extras for plotting (plotly & ggplot2)
```{r setup}
library(popcompr)
library(raster)
library(data.table)
library(foreach)
library(plotly)
library(ggplot2)
library(fasterize)
library(sf)
```

#### Comparing populations at pixel level

You can first get an estimate of the time required to return the comparison:
```{r}
# comparing at pixel level with data included in the package
lesotho_fb_2019 <- raster(system.file("external/lso_facebook_2019.tif", package="popcompr"))
lesotho_wp_2019 <- raster(system.file("external/lso_worldpop_2019.tif", package="popcompr"))

pop_list <- list(lesotho_wp_2019, lesotho_fb_2019)
compare_pop(pop_list, parallel = FALSE, 
                   estimate_time = TRUE)
```

It shouldn't take too long by that estimate, but here we'll work through parallelizing. Here, I use the 
doParallel backend, but other do packages can be used (anything compatible with the `%dopar%` infix in the `foreach` package).

```{r}
# parallelized example
library(doParallel)
cl <- makeCluster(detectCores() - 1) # how many cores do we have available
registerDoParallel(cl)
system.time(
  # defaults to estimate_time = FALSE & resolution of ~ 1km at equator
  exe <- compare_pop(pop_list, parallel = TRUE) 
)
stopCluster(cl)

```

`compare_pop` will warn you if any people were not resampled to the comparison grid (this happens sometimes when pops are at the edge of the original raster).

The output is a raster brick with a layer corresponding to each of the input population rasters. We can vizualize and compare the rasters:

##### Map of differences
```{r}
comp_dt <- data.table(as.data.frame(exe, xy = TRUE)) # we use data.table as it's easier to 

# transform functions for vizualizing difference
transform <- function(x) {
  logged <- log(abs(x) + 1e-6) * sign(x)
  return(logged)
}
inv_trans <- function(x) {
  inv <- (exp(abs(x)) - 1e-6) * sign(x)
  return(round(inv, 2))
}

map_compare <-
  ggplot(comp_dt) +
  geom_raster(aes(x = x, y = y, fill = transform(lso_worldpop_2019 - lso_facebook_2019))) +
  scale_fill_gradient2(labels = inv_trans, 
                       name = "Difference between \n population estimates") +
  coord_quickmap()
map_compare

hex_compare <-
  ggplot(comp_dt) +
  geom_hex(aes(x = lso_worldpop_2019, y = lso_facebook_2019), color = "grey") +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey") +
  scale_fill_distiller(direction = 1, trans = "log",
                       labels = function(x) round(x, -1))
hex_compare
```

You can also make these interactive using plotly/ggplotly functions:
```{r}
ggplotly(map_compare) 
ggplotly(hex_compare) 
```

Check the distribution of differences:
```{r}
ggplot(comp_dt) +
  geom_histogram(aes(x =lso_worldpop_2019 - lso_facebook_2019))
```

### Comparing at the administrative level
You can access country shapefiles using `get_country_shape` which will return a shapefile as an `sf` object. These are available through the [geoBoundaries API](https://www.geoboundaries.org/api.html), where more documentation can be found on available datasets. In addition, `iso_codes` is a dataset provided in the package to find country codes and available admin levels. 
```{r}
# Find the right iso code & see which admin levels are available
dplyr::filter(iso_codes, grepl("Les", country))
```

We get the Lesotho shapefile to the admin level 2 (the `type` argument defaults to a simplified shapefile which is faster for plotting and downloading, but users may prefer unsimplified files for better accuracy see the geoBoundaries API documentation for more information).
```{r}
les_shape <- get_country_shape(country_iso = "LSO", admin_level = 2)
```

Then we can aggregate the population rasters to it:

```{r}
# Get the shapefile at admin 2
les_shape <- aggregate_to_shp(brick = exe, sf = les_shape, max_adjacent = 100)

```
`aggregate_to_shp` will warn you if grid cell values go unallocated to the shapefile. In the case that many people go missing, you can set max_adjacent higher (this determines how many grid cells to buffer to when looking for the nearest non-NA neighbor), but it may also be wise to check the extent & boundaries of the shapefile vs. the population rasters.

We can vizualize these differences as well:
```{r}
map_compare <-
  ggplot(les_shape) +
  geom_sf(aes(fill = transform(lso_worldpop_2019 - lso_facebook_2019))) +
  scale_fill_gradient2(labels = inv_trans, 
                       name = "Difference between \n population estimates")
map_compare

hex_compare <-
  ggplot(les_shape) +
  geom_hex(aes(x = lso_worldpop_2019, y = lso_facebook_2019), color = "grey") +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "grey") +
  scale_fill_distiller(direction = 1, trans = "log",
                       labels = function(x) round(x, -1))
hex_compare
```

You can also make these interactive using plotly/ggplotly functions:
```{r}
ggplotly(map_compare) 
ggplotly(hex_compare) 
```

Check the distribution of differences:
```{r}
ggplot(les_shape) +
  geom_histogram(aes(x =lso_worldpop_2019 - lso_facebook_2019))
```

### User specific inputs
You can use your own raster and shapefile inputs, they just need to be in the WGS84 (lat/long) coordinate system.
